Dynamic Prediction with fPCA 

### 2022-11-02 FDAgroup.Rmd
FDA presentation

library(fcr)
library(face)
library(refund)
library(ggplot2)
library(tidyverse)

@@

- With observations up to $t_m$, predict outcomes (or probabilities of outcome) after that time point
- Prediction updates with new observations

@@

- Goal: to predict future track based on observed track

- For a subject i, we observe a function over t
$$Y_i(t)=f_0(t)+b_i(t)+\epsilon_i(t)$$
- We usually observe $Y_i$ on a series of discrete $t_{ij}$

@@

@@

$$\hat{\boldsymbol{u}} = E(\boldsymbol{u}|\boldsymbol{y}) =  \boldsymbol{\hat{\Gamma}B^T}(\boldsymbol{B\hat{\Gamma}B^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})$$
With the estimated score, we can predict its outcome in following time points

@@

- Predict length-for-age, observed with noise

```{r}
ggplot(df %>% filter(subj%in% 1:10), aes(x=argvals, y=Y, group = subj, col = as.factor(subj)))+
  geom_line(show.legend = F)+
  labs(title = "Length-for-Age for the first 10 subjects",
       x = "time")
```

@@

par(mfrow = c(2, 2))
plot(tnew, fpca_laz$mu.new, type = "l", 
     xlab = "t", ylab = "f0(t)")
plot(tnew, fpca_laz$eigenfunctions[, 1], type = "l", xlab = "t", ylab = "PC1")
plot(tnew, fpca_laz$eigenfunctions[, 2], type = "l", xlab = "t", ylab = "PC2")
plot(tnew, fpca_laz$eigenfunctions[, 3], type = "l", xlab = "t", ylab = "PC3")

@@

@@

- Establish interval prediction
- ? fPCA on pooled prediction, instead of original functions observed with noise
- Extension to exponential family data

>>>>

library(fcr)
library(face)
library(refund)
library(ggplot2)
theme_set(theme_minimal())
library(tidyverse)

@@

- With observations up to $t_m$, predict outcomes (or probabilities of outcome) after that time point
- Prediction updates with new observations
- Mixed model prediction is difficult
  - Not flexible enough for densely measured data
  - Out-of-sample random effects cannot be estimated
- Functional mixed effects model

@@

- Goal: to predict future track based on partially observed track

- For a subject i, we observe a function over t
$$Y_i(t)=f_0(t)+b_i(t)+\epsilon_i(t)$$

@@

- We usually observe $Y_i$ on a series of discrete $t_{ij}$

## Connection to fPCA

@@

$$\hat{\boldsymbol{u}} = E(\boldsymbol{u}|\boldsymbol{y}) =  \boldsymbol{\hat{\Gamma}B^T}(\boldsymbol{B\hat{\Gamma}B^T}+\hat{\sigma_{\epsilon}}^2\boldsymbol{I_m})^{-1}\boldsymbol{y}$$
With the estimated score, we can predict its outcome in following time points

@@

- Predict length-for-age z-score (LAZ), observed with noise
  - Calculated using the age- and sex-specific WHO standard references

```{r}
ggplot(df %>% filter(subj%in% 1:10), aes(x=argvals, y=Y, group = subj, col = as.factor(subj)))+
  geom_line(show.legend = F)+
  labs(title = "Length-for-Age z-score tracks for the 10 subjects",
       x = "time", y ="LAZ")
```

@@

par(mfrow = c(2, 2))
plot(tnew, fpca_laz$mu.new, type = "l", 
     xlab = "t", ylab = "f0(t)")
plot(tnew, fpca_laz$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1")
plot(tnew, fpca_laz$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2")
plot(tnew, fpca_laz$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3")

@@

## Extension to non-Gaussian data

- fPCA on non-Gaussian data is very difficult and computationally intensive
- fPCA on a latent Gaussian variable instead
  - transformation using link function: $g(E(Y))=\boldsymbol{X}^T\boldsymbol{\beta}$
- Get a smooth latent Gaussian function by pooling a series of GLMM
  - Assume we have regularly observed functions $Y_{ij}$, $i = 1... N$ and $j = 1...J$
  - Bin the functional domain into small intervals $t\in {1...T}$
  - Fit a GLMM in each bin to get a latent variables

    $$g(E(Y_{it})) = \beta_{0t}+b_{it}$$
    Let $Z_{it} = g(E(Y_{it}))$, we can do Gaussian fPCA on this latent variable

## Extension to non-Gaussian data

$$Z_i(t) = a_{i1}sin^2(t)+a_{i2}cos^2(t)+a_{i3}t^2+a_{i4}t$$

$$Y_i(t) \sim Binomial(\frac{exp(Z_i(t))}{1+exp(Z_i(t))})$$

## Extension to non-Gaussian data

```{r gen_bi_outcome}
times <- seq(0, 1, by = 0.001)
gen_data <- function(id = 1){
  scores <- rnorm(4)
  pcs <- cbind(sin(times)^2, cos(times)^2, times^2, times)
  Z <- pcs %*% scores
  probs <- exp(Z)/(1+exp(Z))
  Y <- sapply(probs, rbinom, n = 1, size = 1)

  return(data.frame(id = id, argvals = times, Y=Y, Z=Z))
}

df_train2 <- bind_rows(lapply(1:20, gen_data))

ggplot(df_train2 %>% filter(id %in% 1:4))+
  geom_point(aes(x=argvals, y=Y, col = "blue"),  size = 0.1,
            show.legend = F)+
  geom_line(aes(x=argvals, y=Z, col = "Red"), 
            show.legend = F)+
  facet_wrap(~id)+
  labs(title = "Simulated function for the 4 subjects",
       x = "time", y ="")
```

## Extension to non-Gaussian data

```{r, message=FALSE}
# series of mixed models
## bin time
cutbrk <- seq(0, 1, by = 0.01)
cutlabs <- (cutbrk+0.01/2)[1:100]
df_train2$time_bin <- cut(df_train2$argvals, breaks = cutbrk, 
                include.lowest = T,
                labels = cutlabs)
df_train2$time_bin <- as.numeric(as.character(df_train2$time_bin))

# intercept only in each bin mixed models
df_train2_lst <- split(df_train2, f = df_train2$time_bin)
pred_latent <- function(df){
  fit_glm <- lme4::glmer(Y ~ 1 + (1|id), 
                         data = df, family = binomial)
  Zhat <- predict(fit_glm, type = "link")
  return(Zhat)
}

allZhat <- unlist(sapply(df_train2_lst, pred_latent))
df_train_sm <- data.frame(subj = df_train2$id, 
                          argvals = df_train2$time_bin, 
                          y = allZhat) %>% 
  distinct()

ggplot(df_train_sm %>% filter(subj %in% 1:4))+
  geom_line(aes(x=argvals, y=y), 
            show.legend = F)+
  facet_wrap(~subj)+
  labs(title = "Estimated latent function tracks for the 4 subjects",
       x = "time", y ="")

## Extension to non-Gaussian data

```{r}
# fPCA
fpca_Z <- face.sparse(df_train_sm, newdata = df_train_sm, knots = 15,
                        argvals.new = times, calculate.scores = TRUE)

# plot
par(mfrow = c(2, 2))
plot(times, fpca_Z$mu.new, type = "l", 
     xlab = "t", ylab = "f0(t)")
plot(times, fpca_Z$eigenfunctions[, 1], type = "l", xlab = "time", ylab = "PC1")
plot(times, fpca_Z$eigenfunctions[, 2], type = "l", xlab = "time", ylab = "PC2")
plot(times, fpca_Z$eigenfunctions[, 3], type = "l", xlab = "time", ylab = "PC3")

@@

- Simulate a non-Gaussian function and implement methods above
  - Improve estiation of latent functions very well
  - Numeric problems
- Transform latent function back to non-Gaussian function
- Inclusion of covariates
